---
title: "Task 4"
author: "Nzambuli Daniel"
date: "2024-07-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data set Description: You have been provided with monthly sales data for a new retail store over a period of approximately 6.5 years (80 months). The data is measured in thousands of dollars.

```{r}
data = read.csv("timeseries_data.csv")
head(data)
```

```{r}
# convert to excel
writexl::write_xlsx(data, "timeseries_data.xlsx")
```

## 1. Load the data set into RStudio and plot a time series graph to visualize the monthly sales data.

```{r}
library(readxl)
timeseries_data = read_excel("timeseries_data.xlsx")
head(timeseries_data)
```

```{r}
library(tidyr)
library(dplyr)
```

### Provide an interpretation of any observed trends or patterns in the data.

using `ts()` to convert data to timeseries data

-   time is extracted by using `time()`

-   the sales values are extracted using `as.vector()`

```{r}
timeseries_data = timeseries_data %>% separate(Date, into = c("Year", "Month"), sep = "-")
timeseries_data$Year = as.numeric(timeseries_data$Year )
timeseries_data$Month = as.numeric(timeseries_data$Month)
timeseries_data = ts(timeseries_data$Sales, start = c(2017, 1), frequency = 12)
head(timeseries_data)
```

#### Create Time Series Data Frame

```{r}
timeseries_df = data.frame(Time = as.numeric(time(timeseries_data)),
                           Sales = as.vector(timeseries_data))
head(timeseries_df)
```

#### Plot the data

```{r}
library(ggplot2)

ggplot(timeseries_df, aes(x = Time, y = Sales)) +
  geom_line() + 
  labs(title = "Sales Over Time", x = "Time", y = "Sales") +
  theme_minimal()
```

## 2. Perform the Augmented Dickey-Fuller (ADF) test on the sales data to determine if it is stationary

Determine if time series data is stationery; mean, variance, and autocorrelation are constant over time.

$$
\Delta y_t = \alpha + \beta t + \gamma y_{t -1} + \delta_1 \Delta y_{t -1} + \delta_2 \Delta y_{t -2} + ... + + \delta_p \Delta y_{t -p} + \epsilon_t
$$

-   $y_t$​ is the time series.

<!-- -->

-   $\Delta y_t$​ represents the first difference of the series $y_t - y_{t-1}$.

-   $\alpha$ is a constant.

-   $\beta_t$ captures a linear trend.

-   $\gamma$ is the coefficient on $y_{t-1}$​, the lagged value of the series.

-   $\delta_1, \delta_2, ...$ are the coefficients for the lagged differences.

-   $\epsilon_t$ is the error term.

-   $p$ is the number of lagged differences included.

**Hypothesis**

$H_0$ the time series is not stationary

$H_1$ the time series is stationary

using `adf.test` from `tseries` library

### Interpret the results of the test and discuss the implications for further analysis.

```{r}
library(tseries)

k = trunc((nrow(timeseries_df)-1)^(1/3)) # according to r documentation this is the lag value 

adf.test(timeseries_df$Sales, k = k)
```

-   A **more negative** Dickery-Fuller value is evidence against the null hypothesis that the data is not stationary

-   Lag Order – the number of lagged differences used

-   p-value `0.01` reject the $H_0$

**conclude that the data is stationary**

## 3. Apply the Box-Cox transformation to the sales data to stabilize variance.

Normally distribute the data by using $\lambda$ that is varies over real numbers

$$
if\ \lambda \neq 0\:  y(\lambda) = \frac{y^{\lambda} - 1}{\lambda}\\
if\ \lambda = 0 : y(\lambda) = \log {y}
$$

The **log-likelihood** is used to evaluate how well different lambda values normalize the variance and approximate a normal distribution in the transformed data.

### Plot the transformed data and interpret how the transformation impacts the distribution and stationarity of the data.

#### Find the lambda

```{r}
library(MASS)
bx_cox = boxcox(lm(timeseries_df$Sales~1))
lambda = bx_cox$x[which.max(bx_cox$y)]
lambda
```

-   **Zero** is outside the confidence interval of the optimal `lambda`

-   The peak of this curve indicates the lambda value that maximizes the log-likelihood

-   **Vertical Lines** represent the confidence interval for $\lambda$

-   95% represents the critical value for identifying optimum $\lambda$

A $\lambda = 0.7070707$ indicates that a linear model could fit the data adequately while stabilizing variance to make the distribution normal.

The large confidence interval indicates that there is high flexibility in choosing lambda. Because of the flexibility and the $\lambda \approx 1$ applying box-cox transformation may not change the data much.

#### Apply Box-cox

```{r}
timeseries_df$transformed = ((timeseries_df$Sales ^ lambda) - 1)/ lambda
head(timeseries_df)
```

#### Plot

```{r}
plot_data_2 = timeseries_df

ggplot(plot_data_2, aes(x = Time, group = 1)) +
  geom_line(aes(y = Sales, color = "original")) +
  geom_line(aes(y = transformed, color = "transformed")) +
  labs(title = "Time Series and Box-Cox transformed Data of Sales since 2017",
       x = "Year",
       y = "Sales") +
   scale_color_manual(name = "Legend", values = c("original" = "red", "transformed"= "blue"))+
  theme_minimal()
 
```

## 4. Perform the ADF test on the transformed sales data to confirm stationarity

### Interpret the results and compare them with those from Question 2 to assess the effectiveness of the Box-Cox transformation

```{r}
adf.test(timeseries_df$transformed, k = k)
```

-   **Positive Dickey-Fuller** –strong support for $H_0$

-   **P-value** – $\geq 0.05$ fail to reject $H_0$

**Conclude that the new generated data is not stationary**

#### Conclusions

reason

1.  optimum $\lambda$ was selected
2.  The transformation has not achieved stationarity
3.  Because of this a conclusion that the data may have seasonality on top of the trend was made.

The **variance-stabilizing** does not sufficiently achieve stationarity

## 5. Compute and plot the Autocorrelation Function (ACF) for up to 20 lags of the transformed sales data.

### mean of $Y_t$

```{r}
mean_yt = mean(timeseries_df$Sales)
mean_yt
```

### Autocorrelation data

```{r}
auto_cor = timeseries_df[, c(1, 3)]
```

### Auto-corr function

autocorrelation $\rho_k$ for $k = 0, 1, 2, ..., 20$ can be found by

$$
\rho_k = \frac{\sum^{T}_{t = k + 1} (Y_t - \mu) (Y_{t-k} - \mu) }{\sum^{T}_{t = 1} (Y_t - \mu)^2}
$$

```{r}
auto_corr_func = function(data, k){
  n = length(data)
  mu = mean(data)
  
  # T = n 
  # Y_t - mu = y_{k+ 1}, ..., y_{n} like in class (91, 45, 89,...) are moved 1 step below for eack new k the y column. new data mean variation
  # Y_{t -k} - original column but not last value y_1, ..., y_{n-k}. original mean variation
  # denominator is just the original column variance
  num = sum(((data[(k+1): n]) - mu)  *((data[1: (n-k)]) - mu))
  denom = sum((data -mu)^2)
  
  autocorr = num/denom
  
  return(autocorr)
}
```

### fit the auto corr

```{r}
for(i in seq(0,20)){
  col_name = paste("ACF_k_", i)
  auto_cor[[col_name]] = auto_corr_func(auto_cor$transformed, i)
}
```

### check

```{r}
head(auto_cor)
tail(auto_cor)
```

The values in the ACF_k columns are the same so taking one row should be enough for creating the ACF plot.

### Plot

```{r}
library(stats)
plot_data_3 = data.frame(Lag = seq(0,20),
                         ACF = unlist(auto_cor[, 3:ncol(auto_cor)][1,], use.names = FALSE))

ggplot(plot_data_3, aes(x = Lag, y = ACF)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") + 
  geom_hline(yintercept = sd(plot_data_3$ACF), linetype = "dashed", color = "red") + 
  labs(
    title = "Autocorrelation Function (ACF)", 
    x = "Lag", 
    y = "Autocorrelation") +
  theme_minimal()
```

#### Interpret the correlogram to identify any significant autocorrelation patterns and

their implications for forecasting

-   The ACF at lag 0 is always 1, as any series is perfectly correlated with itself at lag 0.

-   Observations close together are closer in correlation, however due to the decrease in autocorrelation from lag 0 the farther two observations are the less the correlation

-   The declining but positive ACF values suggest that the series might be stationary

-   The most recent values give the best predictors

# References

cran.r-project.org. (2020). *Package “tseries” title time series analysis and computational finance*. <https://cran.r-project.org/web/packages/tseries/tseries.pdf>

Jim. (2022, October 23). *Box cox transformation in R \| r-bloggers*. R-Bloggers.com. <https://www.r-bloggers.com/2022/10/box-cox-transformation-in-r/>

RDocumentation. (n.d.). *Adf.test function - rdocumentation*. Www.rdocumentation.org. Retrieved July 13, 2024, from <https://www.rdocumentation.org/packages/tseries/versions/0.10-56/topics/adf.test>

RDocumentation. (n.d.-b). *Ts function \| R documentation*. Www.rdocumentation.org. <https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/ts>
