---
title: "Weighted moving Average"
author: "Nzambuli Daniel"
date: "2024-05-14"
output: html_document
---

# Finding the moving average

```{r}
season = rep(c("summer", "monsoon", "winter", "spring"), 3)

data = data.frame(season = season,
                  year = c(rep(2020, 4),
                           rep(2021, 4),
                           rep(2022, 4)),
                  levels = c(70, 52, 22, 31, 101, 64, 24, 45, 146, 92, 38, 49))

data

```

## moving average

```{r}
library(zoo)
data$moving_avg =  rollapply(data$levels, width = 4, FUN = mean, align = 'right', fill = NA)
data 
```

### ADD Weighted 1

```{r}
wgt = c(1, 2, 3, 4)

weighted_avg = function(x){
  sum(x * wgt)/ sum(wgt)
}

data$wgt_avg = rollapply(data$levels, width = 4, FUN = weighted_avg, align = 'right', fill = NA)

data
```

## Plot the moving and weighted moving average

```{r}
x = seq(1, 9)

plot_data = data.frame(x,
             moving_avg = na.omit(data$moving_avg),
             wgt_mvn_avg = na.omit(data$wgt_avg))

plot_data
```

```{r}
library(ggplot2)

ggplot(plot_data, aes(x=x))+
  geom_line(aes(y = moving_avg), color = "blue")+
  geom_line(aes(y = wgt_mvn_avg), color = "red")
  labs(title = "Line Graphs of Moving Averages",
       x = "X",
       y = "Values") +
  scale_color_manual(name = "Legend", values = c("moving_avg" = "blue", "wgt_mvn_avg" = "red")) +
  theme_minimal()
```

```{r}
ggplot(data, aes(x = seq(1,12)))+
  geom_line(aes(y = moving_avg), color = "blue")+
  geom_line(aes(y = wgt_avg), color = "red")+
  geom_line(aes(y = levels), color = "green")+
  labs(title = "Line Graphs of Moving Averages",
       x = "X",
       y = "Values") +
  scale_color_manual(name = "Legend", values = c("moving_avg" = "blue", "wgt_mvn_avg" = "red","original data" = "green")) +
  theme_minimal()
```

## Exponential Moving Average

```{r}
exp_data = data.frame(year = c(rep(2018, 3),
                               rep(2019, 3),
                               rep(2020, 3),
                               rep(2021, 3)),
                      period = rep(c("I", "II", "III"), 4),
                      no_claims = c(17, 13, 15, 19, 17, 19, 22, 14, 20, 23, 19, 20))

exp_data
```

$$
Exponential\ smoothing = \alpha * y_t + (1-\alpha) * y^{'}_{t-1}
$$

```{r}
exp_smooth = function(alpha, x){
  a = c(x[1])
  
  b = length(a)
  for(i in c(2:length(x))){
    val = (alpha * x[i]) + (1- alpha) * a[i -1]
    a = c(a,val)
  }
  
  return(a)
}
```

```{r}
x_exp = exp_data$no_claims
exp_data$alpha_0.2 = exp_smooth(0.2, x_exp)
exp_data$alpha_0.4 = exp_smooth(0.4, x_exp)
exp_data$alpha_0.6 = exp_smooth(0.6, x_exp)
exp_data$alpha_0.8 = exp_smooth(0.8, x_exp)

exp_data
```

### Plot for exponential

```{r}
ggplot(exp_data, aes(x = seq(1, 12)))+
  geom_line(aes(y = no_claims), color = "blue")+
  geom_line(aes(y = alpha_0.2), color = "red")+
  geom_line(aes(y = alpha_0.4), color = "green")+
  geom_line(aes(y = alpha_0.6), color = "purple")+
  geom_line(aes(y = alpha_0.8), color = "orange")+
  labs(title = "Line Graphs of Moving Averages",
       x = "X",
       y = "Values") +
  scale_color_manual(name = "Legend", values = c("original" = "blue", "alpha 0.2" = "red","alpha 0.4" = "green", "alpha 0.6" = "purple", "alpha 0.8" = "orange")) +
  theme_minimal()
```

## Regression smoothing

Given the data

```{r}
reg_data = data.frame(year = seq(2010, 2018),
             sales = c(52, 54, 48, 60, 61, 66, 70, 80, 92))
```

### Finding $x_t$

```{r}
med_yr = median(reg_data$year)
reg_data$x_t = reg_data$year - med_yr

reg_data
```

### Finding $x_t Y_t$

```{r}
reg_data$x_y = reg_data$x_t * reg_data$x_t
reg_data$x_t_sqrd = reg_data$x_t ^ 2

reg_data
```

```{r}
summary(lm(sales~x_t, data = reg_data))
```

### Extract the intercepts $b_0,\ b_1$

```{r}
lm(sales~x_t, data = reg_data)$coefficients
```

$$
regression smoothing = x_t(data[x_t]) + Intercept
$$

```{r}
reg_data$smooth = 4.80000 * reg_data$x_t + 64.77778 

reg_data
```

### Plot

```{r}
library(ggplot2)

ggplot(aes(x = year), data = reg_data)+
  geom_line(aes(y = sales, color = "original"))+ 
  geom_line(aes(y = smooth, color = "smoothed"))+
  labs(
    title = "regressive smoothing of sales for each year",
    x = "year",
    y = "sales"
  )+
  scale_color_manual(name = "Legend", values = c("original" = "blue", "smoothed" = "red"))+
  theme_minimal()
```

## Quadratic Smoothing

$$
y = \beta_0 + \beta_1t + \beta_2 t^2
$$

```{r}
reg_data$quad_smooth = predict(loess(sales~x_t, reg_data), reg_data)
reg_data
```

```{r}
ggplot(aes(x = year), data = reg_data)+
  geom_line(aes(y = sales, color = "original"))+ 
  geom_line(aes(y = smooth, color = "smoothed"))+
  geom_line(aes(y = quad_smooth, color = "qsmoothed"))+
  labs(
    title = "regressive smoothing of sales for each year",
    x = "year",
    y = "sales"
  )+
  scale_color_manual(name = "Legend", values = c("original" = "blue", "smoothed" = "red", "qsmoothed" = "green"))+
  theme_minimal()
```

## Exponential Trend

$$
zt= log(y)
$$

$$
y_t = \beta_0 e^{\beta_1 * t}\\
log(Y_t) = log(\beta_0) + \beta_1 * t * log(e)\\
= log(\beta_0) + \beta_1t
$$

to find $\beta_0$ we find $e^a$

$$
\sum z_t = na + \beta_1 \sum t 
$$

but $\sum t = 0$

$$
\therefore \frac{\sum z_t}{n} = a = log(\beta_0) 
$$

to find $\beta_1$

$$
\sum t z_t = a \sum t +  \beta_1 \sum t^2\\
remember\ \sum t = 0\\
\therefore \sum t z_t = \beta_1 \sum t^2 
$$

### Exercise

The gross revenue of a company from 2015 to 2022 is given in the following table

| year              | 2015 | 2016 | 2017 | 2018 | 2019 | 2020 | 2021 | 2022 |
|-------------------|------|------|------|------|------|------|------|------|
| **Gross Revenue** | 15   | 45   | 52   | 75   | 106  | 158  | 241  | 314  |

#### i. Fit the exponential trend

```{r}
exp_data = data.frame(year = seq(2015, 2022),
                      revenue = c(15, 45, 52, 75, 106, 158, 241, 314))
exp_data
```

```{r}
exp_data$t = exp_data$year - median(exp_data$year)
exp_data$t_sqrd = exp_data$t^2
exp_data$zt = log(exp_data$revenue)
exp_data
```

**Find** $\beta_0$

```{r}
exp_n = nrow(exp_data)

cat("Beta_0 of the data is: ",exp(sum(exp_data$zt)/ exp_n), "\n")
```

**Find** $beta_1$

```{r}
exp_data$tzt = exp_data$t * exp_data$zt
exp_data
```

```{r}
cat("Beta_1 is given as: ", sum(exp_data$tzt)/ sum(exp_data$t_sqrd), "\n")
```

##### Smooth the values

```{r}
beta_1 = 0.3971438 
beta_0 = 87.17857
exp_data$smoothed = beta_0  * exp(beta_1 * exp_data$t)
exp_data
```

#### ii plot the data

```{r}
library(ggplot2)

ggplot(aes(x = year), data = exp_data)+
  geom_point(aes(y = revenue, color = "original"))+
  geom_line(aes(y = smoothed, color = "smoothed"))+
  labs(
    title = "exponential smoothing",
    x = "year",
    y = "trend"
  )+
  scale_color_manual(name = "Legend", values = c("original" = "red", "smoothed" = "blue"))+
  theme_minimal()
```

#### iii trying exp functions

```{r}
exp_Data = function(df,x, y){
  # transform
  log_data = log(df[[y]])
  
  yr = df[[x]]
  
  # fit 
  model = lm(log_data~yr, data = df )
  
  print(paste("model summary:\n", summary(model)))
  
  # intercepts
  alpha = exp(coef(model)[1])
  beta = coef(model)[2]
  
  cat("The alpha value is and the beta value is:\n\t", alpha, "\t", beta,"\n")
  
  val = df[[x]]
  
  predicted = alpha * exp(beta * val)
  
  return(predicted)
  
  
}
```

```{r}
exp_data$func_smth = exp_Data(exp_data, "t", "revenue")
```

```{r}
exp_data
```
