---
title: "Mid Trimester"
author: "Nzambuli Daniel"
date: "2024-06-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=12, fig.height=6)
```

# The Data

```{r}
library(readxl)
Mid_sem_dataset = read_excel("Mid-sem dataset.xlsx")
Mid_sem_dataset
```

# CASE SCENARIO: TOURIST ANALYSIS IN CAPETOWN

Cape Town, a scenic destination celebrated for its breathtaking landscapes and rich cultural heritage, attracts a diverse group of tourists throughout the year. Its charming old town, scenic hiking trails, vibrant local markets, and tranquil beaches offer visitors a wide array of activities. The attached hypothetical data provides insights into the number of tourists visiting Cape Town from 2019 to 2023.

## 1. Determine the seasonal variation in tourist numbers for each using the ratio-to-moving average approach from 2019 to 2023.

### Quarterly seasonal data only

```{r}
seasonal_data = Mid_sem_dataset[, c(1, 3, 4)]
seasonal_data
```

#### Group by seasons in each year

All the summers in 2019 2020 2022 2023 will be under the label summer ...

```{r}
library(tidyverse)
seasonal_data = seasonal_data %>% group_by(Year, Season) %>% summarise(No_of_tourists = sum(No_of_tourists))
seasonal_data
```

#### Reorder the column of season to appear like the original

target order is `Summer, Autumn, Winter, Spring`

```{r}
seasonal_data$Season = factor(seasonal_data$Season, levels = c("Summer", "Autumn", "Winter", "Spring"))
seasonal_data = seasonal_data[order(seasonal_data$Year, seasonal_data$Season),]
seasonal_data
```

#### Moving average

$$
MA_t = \frac{1}{no.\ seasons} \sum^{t + 3}_{i = 1} x_i\\
= \frac{1}{4}\sum^{t + 3}_{i = 1} x_i
$$

```{r}
mov_avg = zoo::rollmean(seasonal_data$No_of_tourists, k = 4, fill = NA, align = "center")
mov_avg
```

```{r}
seasonal_data$mov_avg = mov_avg
seasonal_data
```

```{r}
rat_mn = zoo::rollmean(seasonal_data$mov_avg, k = 2, fill = NA, align = "center")
rat_mn = c(NA, rat_mn[1:19])
rat_mn
```

#### Centered Moving Average

$$
CMA_t = \frac{1}{2} (x_t + x_{t + 1})
$$

```{r}
seasonal_data$rat_mn = rat_mn
seasonal_data
```

#### Seasonal Ratio

$$
seasonal\ relatives = \frac{actual}{centralized\ avg} * 100
$$

```{r}
seasonal_data$sn_rel = seasonal_data$No_of_tourists/ seasonal_data$rat_mn * 100
seasonal_data
```

#### Pivot the seasonal indexes wider

```{r}
seas_wide = pivot_wider(seasonal_data[, c(1, 2, 6)], names_from = Year, values_from = sn_rel)
seas_wide
```

#### find the median value

```{r}
seas_wide$median = apply(seas_wide[,2:6], 1, function(x) median(x, na.rm = T))
seas_wide
```

#### Adjusted Seasonal indexes

```{r}
tot_seas_median = mean(seas_wide$median)
seas_wide$seas_ind = seas_wide$median / tot_seas_median * 100
seas_wide
```

### Monthly seasonal data

#### Combine the initial code to a function

```{r}
present_month = function(data, x, season_size){
  mov_avg = zoo::rollmean(data[[x]], k = season_size, fill = NA, align = "center")
  print("The moving average is:")
  print(mov_avg)
  rat_mn = zoo::rollmean(mov_avg, k = 2, fill = NA, align = "center")
  rat_mn = c(NA, rat_mn[1:(length(rat_mn) -1)])
  print("The centralized moving average is:")
  print(rat_mn)
  sn_rel = month_data$No_of_tourists/ rat_mn * 100
  print("The seasonal relative is:")
  print(sn_rel)
  long = data.frame(Year = data[,2],
                    Season = data[, 1],
                    sn_rel = sn_rel)
  wide = pivot_wider(long, names_from = Year, values_from = sn_rel)
  print("The data in wide format is:")
  print(wide)
  wide$median = apply(wide[,2:ncol(wide)], 1, function(x) median(x, na.rm = T))
  tot_seas_median = mean(wide$median)
  wide$seas_ind = wide$median / tot_seas_median * 100
  print("With seasonal index")
  return(wide)
}
```

#### The annual data

```{r}
month_data = Mid_sem_dataset[, c(2, 3, 4)]
colnames(month_data)
```

```{r}
month_wide = present_month(month_data,"No_of_tourists", 12)
month_wide
```

## 2. Obtain a clearer view of the underlying trend by eliminating seasonal fluctuations.

### Monthly Data

```{r}
monthly_smoothed_data = data.frame(month = rep(month_wide$Month, 5),
                               year = rep(seq(2019, 2023), 12),
                               tourists = month_data$No_of_tourists,
                               adj_seas_ind = rep(month_wide$seas_ind, 5))

monthly_smoothed_data$deseason = monthly_smoothed_data$tourists/ monthly_smoothed_data$adj_seas_ind * 100
monthly_smoothed_data
```

### Quarterly data

```{r}
quarter_smoothed_data = data.frame(quarter = rep(seas_wide$Season, 5),
                               year = rep(seq(2019, 2023), 4),
                               adj_seas_ind = rep(seas_wide$seas_ind, 5))
quarter_smoothed_data = quarter_smoothed_data[order(quarter_smoothed_data$year),]
quarter_smoothed_data$tourist = seasonal_data$No_of_tourists


quarter_smoothed_data$deseasoned = quarter_smoothed_data$tourist/quarter_smoothed_data$adj_seas_ind * 100

rownames(quarter_smoothed_data) = NULL
quarter_smoothed_data

```

## 3. Compute the long-term trend in tourist numbers over the specified period by applying the suitable trend analysis to the deseasonalized data.

if a linear model is to be used

$$
trend\ line = \beta_0 + \beta_1 t
$$

otherwise for quadratic

$$
trend\ line = \beta_0 + \beta_1 t + \beta_2t^2
$$

and for exponential

$$
trend\ line = \beta_0 + e^{\beta_1 t}
$$

### Generate t

```{r}
nrow(monthly_smoothed_data)
```

60 is even so

$$
x_t = 2(t - median(t))
$$

### Month data

```{r}
monthly_smoothed_data$t = seq(1:nrow(monthly_smoothed_data))
med_month_t = median(monthly_smoothed_data$t)
monthly_smoothed_data$x_t = 2*(monthly_smoothed_data$t - med_month_t)
monthly_smoothed_data$t_sqrd = monthly_smoothed_data$t^2
monthly_smoothed_data

```

After the transformation, the normal equations for linear trend line are:

$$
\sum Y_t = n\beta_0 + \beta_1\sum X_t 
$$

```{r}
sum(monthly_smoothed_data$x_t)
```

therefore

$$
\frac{\sum Y_t}{n} = \beta_0
$$

additionally

$$
\sum X_t Y_t = \beta_0 \sum X_t + \beta_1 \sum X_t^2\\
=  \beta_1 \sum X_t^2
$$

add the column for $X_t Y_t$ and $X_t^2$

```{r}
monthly_smoothed_data$XY = monthly_smoothed_data$deseason * monthly_smoothed_data$x_t
monthly_smoothed_data$x_sqrd = monthly_smoothed_data$x_t^2
monthly_smoothed_data
```

### Trend Values

#### Test the best trend

#### 1. simple linear

```{r}
summary(lm(deseason~x_t, data = monthly_smoothed_data))
```

#### 2. quadratic trend

```{r}
summary(lm(deseason~poly(x_t, 2, raw = T), data = monthly_smoothed_data))
```

```{r}
summary(lm(deseason~x_t + x_sqrd, data = monthly_smoothed_data))
```

#### 3. exponential

```{r}
summary(nls(deseason~a + exp(b * x_t), data = monthly_smoothed_data, start = list(a = 1, b = 0.1)))
```

#### compare the models

Lower **Residual R** values indicate a better fit.

```{r}
quadratic_model = lm(deseason~poly(x_t, 2, raw = T), data = monthly_smoothed_data)
linear_model = lm(deseason~x_t, data = monthly_smoothed_data)

linear_residual_R = summary(linear_model)$sigma
quadrat_residual_R = summary(quadratic_model)$sigma
if (linear_residual_R < quadrat_residual_R){
  print("Linear Model is the best model")
}else if(linear_residual_R == quadrat_residual_R){
  print("any model can be used")
}else{
  print("Quadratic Model is the best model")
}


```

#### quad vs exp

```{r}
exp_model = nls(deseason~a + exp(b * x_t), data = monthly_smoothed_data, start = list(a = 1, b = 0.1))

exp_residual_R = summary(exp_model)$sigma
if (exp_residual_R < quadrat_residual_R){
  print("Exponential Model is the best model")
}else if(linear_residual_R == quadrat_residual_R){
  print("any model can be used; Exponential or Quadratic")
}else{
  print("Quadratic Model is the best model")
}

```

#### Choosing quadratic model

```{r}
coeff_monthly = coef(quadratic_model)
coeff_monthly
monthly_b0 = coeff_monthly[1]
monthly_b1 = coeff_monthly[2]
monthly_b2 = coeff_monthly[3]
```

```{r}
monthly_smoothed_data$trend = monthly_b0 + (monthly_b1 * monthly_smoothed_data$t) + (monthly_b2 * monthly_smoothed_data$t_sqrd)
monthly_smoothed_data
```

### quarter

```{r}
nrow(quarter_smoothed_data)
```

```{r}
quarter_smoothed_data$t = seq(1, 20)
med_quarter = median(quarter_smoothed_data$t)
quarter_smoothed_data$x = 2 * (quarter_smoothed_data$t- med_quarter)
```

#### Compare the models

```{r}
quadratic_model_q = lm(deseasoned~poly(x, 2, raw = T), data = quarter_smoothed_data)
linear_model_q = lm(deseasoned~x, data = quarter_smoothed_data)

linear_residual_R = summary(linear_model_q)$sigma
quadrat_residual_R = summary(quadratic_model_q)$sigma
if (linear_residual_R < quadrat_residual_R){
  print("Linear Model is the best model")
}else if(linear_residual_R == quadrat_residual_R){
  print("any model can be used")
}else{
  print("Quadratic Model is the best model")
}
```

```{r}
quarter_smoothed_data$log_deseason = log(quarter_smoothed_data$deseasoned)
exp_model_q = lm(log_deseason~x, data = quarter_smoothed_data)

exp_residual_R = summary(exp_model_q)$sigma
if (exp_residual_R < quadrat_residual_R){
  print("Exponential Model is the best model")
}else if(linear_residual_R == quadrat_residual_R){
  print("any model can be used; Exponential or Quadratic")
}else{
  print("Quadratic Model is the best model")
}

```

#### Generate the rest of the columns

```{r}
quarter_smoothed_data$XY = quarter_smoothed_data$deseasoned * quarter_smoothed_data$x
quarter_smoothed_data$x_sqrd = quarter_smoothed_data$x^2
quarter_smoothed_data$t_sqrd = quarter_smoothed_data$t^2
quarter_coeff = coef(exp_model_q)
quarter_coeff
quarter_b0 = exp(quarter_coeff[1])
quarter_b0
quarter_b1 = quarter_coeff[2]
quarter_b1
quarter_smoothed_data$trend = quarter_b0 + exp(quarter_b1 * quarter_smoothed_data$t)
quarter_smoothed_data
```

## 4. Obtain the cyclic index to better understand the cyclical variations in tourist numbers that may be influenced by economic cycles or other factors.

$$
cyclic\ index = \frac{deseasonalized\ data}{trend\ value} * 100
$$

### Monthly

```{r}
monthly_smoothed_data$cyclic = monthly_smoothed_data$deseason/ monthly_smoothed_data$trend * 100
monthly_smoothed_data
```

### Quarters

```{r}
quarter_smoothed_data$cyclic = quarter_smoothed_data$deseason/ quarter_smoothed_data$trend * 100
quarter_smoothed_data
```

## 5. Generate graphs of the data to illustrate the seasonal, trend, and cyclic components in the number of tourists visiting Cape Town.

### Monthly

```{r}
months = c(
  "January", "February", "March", "April", "May", "June",
  "July", "August", "September", "October", "November", "December"
)

monthly_smoothed_data$month = factor(monthly_smoothed_data$month, levels = months)
monthly_smoothed_data$season_year = paste(monthly_smoothed_data$month, monthly_smoothed_data$year)
monthly_smoothed_data$season_year = factor(monthly_smoothed_data$season_year, levels = unique(monthly_smoothed_data$season_year))
# 
# 
ggplot(monthly_smoothed_data, aes(x = season_year)) +
  geom_line(aes(y = tourists, color = "original"), group = 1) +
  # geom_line(aes(y = adj_seas_ind, color = "seasonal"), group = 1) +
  geom_line(aes(y = deseason, color = "deseason"), group = 1, linetype = "dotted") +
  geom_line(aes(y = trend, color = "trend"), group = 1) +
  geom_line(aes(y = cyclic, color = "cyclic"), group = 1) +
  labs(title = "Seasonal Index from Summer 2019 to Spring 2023",
       x = "Time",
       y = "Seasonal Index") +
  theme_minimal() +
  scale_color_manual(name = "Legend", values = c("original" = "black", "trend"= "blue", "cyclic" = "darkgreen", "deseason" = "red"))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

> Using the monthly data,
>
> 1.  There is an upward trend in the visitors to Capetown
> 2.  There does not seem to be a cycle where visitors stop or increase when visiting cape town
> 3.  There is a very low rise in visitors to Capetown up to 2021 where there seems to be a spike and then a drop after may 2021

### Quarter

```{r}
quarter_smoothed_data$date = as.Date(paste(quarter_smoothed_data$year, as.numeric(quarter_smoothed_data$quarter) * 3 - 2, "1", sep = "-"), format = "%Y-%m-%d")

quarter_smoothed_data$season_year = paste(quarter_smoothed_data$quarter, quarter_smoothed_data$year)

ggplot(quarter_smoothed_data, aes(x = date)) +
  geom_line(aes(y = tourist, color = "original"), group = 1) +
  geom_line(aes(y = deseasoned, color = "deseason"), group = 1, linetype = "dotted") +
  geom_line(aes(y = trend, color = "trend"), group = 1) +
  geom_line(aes(y = cyclic, color = "cyclic"), group = 1) +
  labs(title = "Quarterly Seasonal Index from Summer 2019 to Spring 2023",
       x = "Time",
       y = "Seasonal Index") +
  theme_minimal() +
  scale_color_manual(name = "Legend", values = c("original" = "black", "trend"= "blue", "cyclic" = "darkgreen", "deseason" = "purple"))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

> The quarter trend
>
> 1.  There seems to be no trend when viewed across quarters
> 2.  There are peaks only during summer and then the visitation drops.
