---
title: "week 5 take home"
author: "Nzambuli Daniel"
date: "2024-06-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question

The marketing manager of a company that manufactures and distributes farming equipment (harvesting, plough and tractors) recorded the number of farming units sold quarterly for a period 2018 to 2020 which are given in the following table

| Quarter |  Q1 |  Q2 |  Q3 |
|:-------:|----:|----:|----:|
|  2018   |  48 |  41 |  60 |
|  2019   |  58 |  52 |  68 |
|  2020   |  60 |  58 |  75 |

## Find the quarterly seasonal indexes for farming equipment sold using the ratio to trend method by assuming that there is no cyclic effect

```{r}
sales_data = data.frame(
  Year = c(2018, 2019, 2020),
  Q1 = c(48, 58, 60),
  Q2 = c(41, 52, 58),
  Q3 = c(60, 68, 75),
  Q4 = c(65, 74, 78)
)

sales_data
```

```{r}
library(tidyr)
sales_long = pivot_longer(sales_data, cols = starts_with("Q"), names_to = "Quarter", values_to = "Sales")
sales_long
```

```{r}
sales_data$annual_means = rowMeans(sales_data[,2:ncol(sales_data)])
sales_data
```

### Predicting $\beta_0$ and $\beta_1$

```{r}
predictors = coef(lm(Year~annual_means, data = sales_data))
predictors
```

## Do Seasonal forces significantly influence the sale of farming equipment? Comment
